---
title: "Simulation 7: Stroop Task Switching"
author:
  - name: Matthew J. C. Crump
    orcid: 0000-0002-5612-0090
    email: mcrump@brooklyn.cuny.edu
    affiliations:
      - name: Brooklyn College of CUNY
abstract: "This simulation involves a Stroop task and task-switching. The instruction for a given trial will be to identify the word or color."
date: 6-28-2023
title-block-style: default
order: 8
format:
  html:
    code-overflow: wrap
    code-fold: true
    code-summary: "Show the code"
---

## State:

Started. Finished.


## Recap

This simulation continues in the direction of determining whether GPT will produce patterns of data typical of human performance in a Stroop task, even for conditions and task demands that are not explicitly prompted. 

Simulation 1 showed that GPT appears to generate data that contain congruency sequence effects. Simulation 2 showed that GPT may generate data sensitive to a list-wide proportion congruent manipulation. 

This simulation looks at task-switching performance. In a Stroop task it is possible to have participants name the ink-color or the word. Word-naming is generally faster than color naming. People show slower RTs and larger Stroop effects for the color naming task, and overall faster RTs and much smaller Stroop effects for the word naming task. If the instructions are intermixed within a block of trials, then people also show a task-switching effect. Responses are faster when the task repeats across trials, compared to when the task switches across trials. 

## Load libraries

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(openai)
library(patchwork)
library(xtable)
```


## Run model

Notes: 15 simulated subjects. 48 Stroop trials each. 50/50 congruent and incongruent trials. Half of the trials are color-naming, the other half are word-naming, randomly intermixed.

The prompt does no longer mentions the Stroop effect, instead it mentions word or color naming task.

Used gpt-3.5-turbo-16k, with max tokens 10000.

**Problems:** Still getting the occasional invalid JSON back, mostly due to the chatbot prefacing its response with a message before the JSON. This thread may be helpful <https://community.openai.com/t/getting-response-data-as-a-fixed-consistent-json-response/28471/31?page=2>.


```{r, eval=FALSE}
# use the colors red, green, blue, and yellow

# four possible congruent items
congruent_items <- data.frame(word  = c("red","green","blue","yellow"),
                              color = c("red","green","blue","yellow"))

# 12 possible congruent items
incongruent_items <- data.frame(word  = c("red","red","red",
                                          "green","green","green",
                                          "blue","blue","blue",
                                          "yellow","yellow","yellow"),
                              color = c("green","blue","yellow",
                                        "blue","yellow","red",
                                        "red","yellow","green",
                                        "red","blue","green"))


#set up variables to store data
all_sim_data <- tibble()
gpt_response_list <- list()

# request multiple subjects
# submit a query to open ai using the following prompt
# note: responses in JSON format are requested

for(i in 1:15){
  print(i)
  
  # construct trials data frame
  congruent_trials <- congruent_items[rep(1:nrow(congruent_items),3),]
  incongruent_trials <- incongruent_items[rep(1:nrow(incongruent_items),1),]

  trials <- rbind(congruent_trials,
                  incongruent_trials,
                  congruent_trials,
                  incongruent_trials
                  ) %>%
    mutate(instruction = rep(c("Identify color","Identify word"),each=24))
  
  trials <- trials[sample(1:nrow(trials)),]
  trials <- trials %>%
    mutate(trial = 1:nrow(trials),
           response = "?",
           reaction_time = "?") %>%
    relocate(instruction) %>%
    relocate(trial)
  
   # run the api call to openai
  
 gpt_response <- create_chat_completion(
   model = "gpt-3.5-turbo-16k",
   max_tokens = 10000,
   messages = list(
       list(
           "role" = "system",
           "content" = "You are a simulated participant in a human cognition experiment. Complete the task as instructed and record your simulated responses in a JSON file"),
       list("role" = "assistant",
            "content" = "OK, I am ready."),
       list("role" = "user",
           "content" = paste('You are a simulated participant in a human cognition experiment. Complete the task as instructed and record your simulated responses in a JSON file. Your task is to simulate human performance in a word and color naming task. You will be the task in the form a JSON object. The JSON object contains the word and color presented on each trial. Your task is to read the task instruction for each trial. If the instruction is to name the color, then identify the color as quickly and accurately as a human would. If the instruction is to name the word, then identify the word as quickly and accurately as a human would. The JSON object contains the symbol ? in locations where you will generate simulated responses. You will generate a simulated identification response, and a simulated reaction time for each trial. Put the simulated identification response and reaction time into a JSON array using this format: [{"trial": "trial number, integer", "instruction" = "the task instruction, string", "word": "the name of the word, string","color": "the color of the word, string","response": "the simulated identification response, string","reaction_time": "the simulated reaction time, milliseconds an integer"}].',
                       "\n\n",
                       jsonlite::toJSON(trials), collapse = "\n")
           
       )
   )
)
  
  # save the output from openai
  gpt_response_list[[i]] <- gpt_response
  print(gpt_response$usage$total_tokens)
  
  # validate the JSON  
  test_JSON <- jsonlite::validate(gpt_response$choices$message.content)
  print(test_JSON)

  # validation checks pass, write the simulated data to all_sim_data 
  if(test_JSON == TRUE){
    sim_data <- jsonlite::fromJSON(gpt_response$choices$message.content)
    
    if(sum(names(sim_data) == c("trial","instruction","word","color","response","reaction_time")) == 6) {
      
      sim_data <- sim_data %>%
        mutate(sim_subject = i)
  
      all_sim_data <- rbind(all_sim_data,sim_data)
    }
    
  }
}

# model responses are in JSON format
save.image("data/simulation_7.RData")
```



## Analysis

```{r}
load(file = "data/simulation_7.RData")
```

The LLM occasionally returns invalid JSON. The simulation ran 15 times.

```{r}
total_subjects <- length(unique(all_sim_data$sim_subject))
```

There were `r total_subjects` out of 15 valid simulated subjects. 

### Reaction time analysis


```{r}
all_sim_data <- all_sim_data %>%
  mutate(reaction_time = as.numeric(reaction_time))

# get mean RTs in each condition for each subject
rt_data_subject_congruency <- all_sim_data %>%
  mutate(congruency = case_when(word == color ~ "congruent",
                                word != color ~ "incongruent")) %>%
  mutate(accuracy = case_when(instruction == "Identify color" & response == color ~ TRUE,
                              instruction == "Identify color" & response != color ~ FALSE,
                              instruction == "Identify word" & response == word ~ TRUE,
                              instruction == "Identify word" & response != word ~ FALSE)) %>%
  filter(accuracy == TRUE) %>%
  group_by(instruction,congruency,sim_subject) %>%
  summarize(mean_rt = mean(reaction_time), .groups = "drop")

# make plots

ggplot(rt_data_subject_congruency, aes(x = congruency,
                                       y = mean_rt))+
  geom_violin()+
  stat_summary(fun = "mean",
               geom = "crossbar",
               color = "red")+
  geom_point()+
  theme_classic(base_size=15)+
  ylab("Mean Simulated Reaction Time") +
  facet_wrap(~instruction)

```
The figure shows similarly sized Stroop effects for the color naming and word naming tasks. This is not the kind of pattern that would be expected from human participants.

### A closer look at reaction times

This time the reaction times look like they came from a normal distribution.


```{r}
ggplot(all_sim_data, aes(x=reaction_time))+
  geom_histogram(binwidth=50, color="white")+
  theme_classic()+
  xlab("Simulated Reaction Times")
```

The prompt did not specify to produce values with different endings. As with previous simulations, the model prefers values ending in 0.

```{r}
all_sim_data <- all_sim_data %>%
  mutate(ending_digit = stringr::str_extract(all_sim_data$reaction_time, "\\d$")) %>%
  mutate(ending_digit = as.numeric(ending_digit))

ggplot(all_sim_data, aes(x=ending_digit))+
  geom_histogram(binwidth=1, color="white")+
  scale_x_continuous(breaks=seq(0,9,1))+
  theme_classic(base_size = 10)+
  xlab("Simulated RT Ones Digit")
```

## Accuracy Analysis

The model performs perfectly on congruent trials, and sometimes imperfectly on incongruent trials.


```{r}
# report accuracy data
accuracy_data_subject <- all_sim_data %>%
  mutate(congruency = case_when(word == color ~ "congruent",
                                word != color ~ "incongruent")) %>%
  mutate(accuracy = case_when(instruction == "Identify color" & response == color ~ TRUE,
                              instruction == "Identify color" & response != color ~ FALSE,
                              instruction == "Identify word" & response == word ~ TRUE,
                              instruction == "Identify word" & response != word ~ FALSE)) %>%
  group_by(instruction,congruency,sim_subject) %>%
  summarize(proportion_correct = mean(accuracy), .groups = "drop")

ggplot(accuracy_data_subject, aes(x = congruency,
                                  y = proportion_correct))+
  geom_violin()+
  stat_summary(fun = "mean",
               geom = "crossbar",
               color = "red")+
  geom_point()+
  theme_classic(base_size=15)+
  ylab("Simulated Proportion Correct")+
  xlab("Congruency") +
  facet_wrap(~instruction)

```

## Task-switching

```{r}
# add last trial congruency as a factor
all_sim_data$last_trial_task <- c(NA,all_sim_data$instruction[1:(dim(all_sim_data)[1]-1)])

all_sim_data <- all_sim_data %>%
  mutate(last_trial_task = case_when(trial == 1 ~ NA,
                                          trial != 1 ~ last_trial_task)) %>%
  mutate(switch_type = case_when(instruction == last_trial_task ~ "repeat",
                                 instruction != last_trial_task ~ "switch")
         )

# get mean RTs in each condition for each subject
rt_data_subject_switch <- all_sim_data %>%
  mutate(congruency = case_when(word == color ~ "congruent",
                                word != color ~ "incongruent")) %>%
  mutate(accuracy = case_when(instruction == "Identify color" & response == color ~ TRUE,
                              instruction == "Identify color" & response != color ~ FALSE,
                              instruction == "Identify word" & response == word ~ TRUE,
                              instruction == "Identify word" & response != word ~ FALSE)) %>%
  filter(accuracy == TRUE,
         is.na(last_trial_task) == FALSE) %>%
  group_by(switch_type,instruction,sim_subject) %>%
  summarize(mean_rt = mean(reaction_time), .groups = "drop")

# make plots

ggplot(rt_data_subject_switch, aes(x = switch_type,
                                       y = mean_rt))+
  geom_violin()+
  stat_summary(fun = "mean",
               geom = "crossbar",
               color = "red")+
  geom_point()+
  theme_classic(base_size=15)+
  ylab("Mean Simulated Reaction Time") +
  facet_wrap(~instruction)
```

The model did not generate faster simulated reaction times for repeat than switch trials, which is commonly found in human data. 

```{r}
# report accuracy data
accuracy_data_subject <- all_sim_data %>%
  mutate(congruency = case_when(word == color ~ "congruent",
                                word != color ~ "incongruent")) %>%
  mutate(accuracy = case_when(instruction == "Identify color" & response == color ~ TRUE,
                              instruction == "Identify color" & response != color ~ FALSE,
                              instruction == "Identify word" & response == word ~ TRUE,
                              instruction == "Identify word" & response != word ~ FALSE)) %>%
  filter(is.na(last_trial_task) == FALSE) %>%
  group_by(switch_type,instruction,sim_subject) %>%
  summarize(proportion_correct = mean(accuracy), .groups = "drop")

ggplot(accuracy_data_subject, aes(x = switch_type,
                                  y = proportion_correct))+
  geom_violin()+
  stat_summary(fun = "mean",
               geom = "crossbar",
               color = "red")+
  geom_point()+
  theme_classic(base_size=15)+
  ylab("Simulated Proportion Correct")+
  xlab("Congruency") +
  facet_wrap(~instruction)
```
Accuracy was worse for the color identification instructions than the word identification instructions. But, there was no switch effect on accuracy either. 


## Discussion

This simulation asked whether GPT would return ancillary effects that conform to typical patterns in a Stroop, especially when those results are not explicitly prompted. The simulation was given a Stroop task, and instructed to name the word or color on each trial. Human participants would typically show faster word naming than color naming responses, and smaller Stroop effects on word than color naming trials. The model did not produce faster word than color naming responses, nor did it generate smaller Stroop effects for word than color naming trials. Humans typically show faster responses on task-repeat than task-switch trials. The model did not show the task switching effect.














